### 2015-10-12 paper-1/main.tex
Changes based on Dan's feedback.

This paper introduces a new discrete distribution suggested by curtailed 
sampling rules common in early-stage clinical trials. We derive the 
distribution of the smallest number of independent Bernoulli($p$) trials 
needed in order to observe either $s$ successes or $t$ failures. The closed 
form expression for the distribution as well as the compound distribution are 
derived. Properties of the distribution are shown and discussed. A case study 
is presented showing how the distribution can be used to monitor sequential 
enrollment of clinical trials with binary outcomes as well as providing 
post-hoc analysis of completed trials.

@@

Consider the design for a Phase II clinical trial to investigated the 
efficacy of iniparib in patients with breast cancer gene-associated (BRCA) 
ovarian cancer [\ref{Sanofi 2013}]. In the first stage, 12 patients were 
treated with iniparib (BSI-201/SAR240550). The endpoint was defined by 
Response Evaluation Criteria in Solid Tumor (RECIST). If at least two of
these patients respond favorably, then the trial will proceed to the next 
stage where additional patients are able to be treated. If fewer than two 
respond then the trial would be terminated.

The maximum sample size is 12 but the number of patients necessary to 
reach and endpoint can be less. Our goal is to describe the distribution of 
the enrollment size of the design for 
planning purposes and to use this distribution to estimate the
success probability.  If all 12 patients are enrolled at once, 
as in the classic 
design, then the sample size is 12. However, in most clinical trials, the 
patients are enrolled sequentially, often with one patient's outcome realized 
before the next one enters the trial. In the present example, observing two 
successful patients allows us to proceed to the next stage. Similarly 11 
observed treatment failures also ends the stage. This sampling mechanism, in 
which the experiment ends as soon as any of the endpoints is reached, is 
call {\em curtailed sampling}. Under curtailed sampling the range of the 
sample size is between two and 12.

We assume each of the patient outcomes can be modeled as independent, 
identically distributed Bernoulli($p$) random variables. The realization of 
the trial is a sequence of these random variables that stops when either a 
specified number of success or failures is reached. As in the 
previous example, two successes were reached after enrolling 10 
patients (one in the third step and one at the $10^{th}$). The trial is 
visualized in Fig.~\ref{fig:kane_viz}. The vertical axis denotes the number of 
successful outcomes. The horizontal axis denotes the number of patients that 
have been enrolled. The horizontal and vertical line segments represent 
boundary endpoints for the trial.

@@

Let $\,b_1, b_2, \ldots \,$ denote a sequence of independent, identically 
distributed, Bernoulli random variables with $\mathbb{P}\{b_i=1\}=p$, for 
probability parameter $0\leq p \leq 1$. In the clinical trial setting 
$\,b_i = 0$ corresponds to an unsuccessful outcome.  Let $s$ and $t$ be 
positive integers.  Define the stopped negative binomial (SNB) random 
variable $Y$ as the smallest 
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either} 
$\,s\,$ successes {\em or} $\,t\,$ failures. The distribution of $\,Y\,$ has 
support on integer values in the range 
\begin{equation*}                                     %   (3)
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}
\end{equation*}
and it is distributed as
\begin{equation} \label{eqn:pmf}
\mathbb{P}[Y=k] = S(k, p, s)\left[s \leq k\right] + 
  S(k, 1-p, t) \left[ t \leq k \right]
\end{equation}
where 
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
$S$ is the negative binomial probability truncated to a subset of the support 
- in this case, $s$ and $t$ respectively.
From Eqn.~\ref{eqn:pmf} one can see that the mass at 
any point on the support of the distribution may come from either 
$X_k = s$, captured in the function $S$, or $X_k = k-t$, captured in the 
function $T$. The support of 
$S$ is defined by $s \leq S \leq s+t-1$ and the support of $T$ is defined 
by $t \leq T \leq s+t-1$.

@@

To show the result in Eqn.~\ref{eqn:pmf} consider the 
process $\mathbf{X} = \left[X(k) : k \in \mathbb{Z}_{\ge 0} \right]$ 
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} \left[ k-t < X_k < s \right].
\end{equation*}
As before the $b_{k+1}$'s are distributed as Bernoulli($p$) for all $k$ in the 
domain of $\mathbf{X}$. The process can be conceptualized as a series of coin 
flips that stops when either $s$ heads or $t$ tails are reached. At each step 
a coin is flipped. If it is heads, the process advances one diagonally in the 
positive horizontal and vertical direction. Otherwise, it advances in the 
positive horizontal direction only. When the process stops
either $X_k = s$ or $X_k = k-t$. One example is given in 
Fig.~\ref{fig:kane_viz}.

\begin{prop}
The distribution of the stopping time $\argminl_k [X_k \geq s, X_k \leq k-t]$ 
is equal to the SNB and is given in (\ref{eqn:pmf}). 
\end{prop}
\begin{proof}
%The proof will proceed in two parts. First, a combinatorial justification 
%will be given for the probability mass value on each element of the support. 
%Second, it will be shown that the sum of the masses over the support sums to 
%one.

The probability that a given realization of $\mathbf{X}$ reaches $s$ at 
the $k$th outcome is the probability that, at time $k-1$ there are $s-1$ 
successful outcomes and $k-s$ unsuccessful outcomes multiplied by 
the probability of a 
success at time $k$ (\ref{eqn:N}). The probability $X_k = k-t$ 
is the probability that, at outcome $k-1$ there are $k-t$ successful outcomes 
and $t-1$ unsuccessful outcomes multiplied by the probability of an 
unsuccessful outcome at time $k$.  Finally, we need to show that 
the sum
\begin{equation} \label{eqn:sum_proof}
R = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{equation}
is equal to one.
First, substitute $i=k-s$ in the first summation and
$j=k-t$ in the second. 
\begin{equation*}
{j+s-1 \choose s-1} = {j+s-1 \choose j}
\end{equation*}
Then %(\ref{eqn:sum_proof}) becomes:
%\begin{equation*}
%R = \sum_{i=0}^{t-1} {i+s-1 \choose s-1} p^s (1-p)^i +
%\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
%\end{equation*}
$R$ can be written
as the cumulative distribution function of two 
negative binomial distributions
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

Let $I_p(s, t)$ be the {\em regularized incomplete beta function}, which is 
also one minus the c.d.f. of the negative binomial distribution. This
function satisfies $I_p(s, t) = 1-I_{1-p}(t, s)$.  Then it can be proven
that \ref{eqn:pmf} is a valid probability mass by showing that the
probability mass function sums to one.
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-I_p(s, t) + 1 - I_{1-p}(t, s) \\
   &= 1. 
\end{align*}
\end{proof}

\begin{prop} Let $S$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $S$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xS} = \frac{p e^x}{1 - qe^x} I_{qe^x} (s, t) + \frac{qe^x}{1-pe^x} I_{pe^x}(s, t)
\end{equation}
for $q = 1-p$ when $x \leq -\log(q)$ and $x \leq -\log(p)$.
\end{prop}
\begin{proof}
By definition, the MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} q^t e^{kx}
\end{equation*}
and this can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose k-t} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
Taking the first summation in Equation \ref{eqn:first_sum}:
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s) \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s I_{qe^x}(s, t).
\end{align*}
Since the incomplete beta function has support on zero to one 
$qe^x \leq 1$. This implies that $x \leq -\log(q)$.

@@

Let us assume that $p$ has a Beta distribution, with constant prior 
parameters $\alpha$ and $\beta$. Then the closed form 
posterior distribution of the SNB is
\begin{align}
\mathbb{P} \left[Y = k | \mathbf{b}, \alpha, \beta \right] &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
  [s \leq k \leq s+k-1] + \nonumber \\
  & {k-1 \choose k-t} 
  \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
  [t \leq k \leq s+k-1]
\end{align}
where $B$ is the Beta function and $\mathbf{b}$ is the sequence of Bernoulli data.

@@

\begin{prop}
The posterior PMF of the Stopped Negative Binomial distribution with a Beta($\alpha$, $\beta$) prior is:
\begin{align} \label{eqn:posterior}
f(k | s, t, \alpha, \beta) &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)}     [s \leq k \leq s+k-1] + \nonumber \\
  & {k-1 \choose k-t} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    [t \leq k \leq s+k-1]
\end{align}
where $B$ is the Beta function.
\end{prop}
\begin{proof}
For notational simplicity, assume that $s,t \leq k \leq s+t-1$. When this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose k-t} p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose k-t} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp
\end{align*}
The result follows by applying the definition of the Beta function to the integral terms.
\end{proof}

>>>>

This paper introduces a new discrete distribution suggested by curtailed 
sampling rules common in early-stage clinical trials. We derive the 
distribution of the smallest number of independent and identically 
distributed Bernoulli trials 
needed in order to observe either $s$ successes or $t$ failures. The closed 
form expression for the distribution as well as the compound distribution are 
derived. Properties of the distribution are shown and discussed. A case study 
is presented showing how the distribution can be used to monitor sequential 
enrollment of clinical trials with binary outcomes as well as providing 
post-hoc analysis of completed trials.

@@

Consider the design for a Phase II clinical trial that investigates the 
efficacy of iniparib in patients with breast cancer gene-associated (BRCA) 
ovarian cancer [\ref{Sanofi 2013}]. In the first stage, 12 patients were 
treated with iniparib (BSI-201/SAR240550). The outcome was defined by 
Response Evaluation Criteria in Solid Tumor (RECIST). There were two 
endpoints: terminate or enroll additional patients at the end of this 
stage. If at least two of
these patients respond favorably, then the trial will proceed to the next 
stage where additional patients were able to be treated. If fewer than two 
respond then the trial would have been terminated.

The maximum sample size was 12 but the number of patients necessary to 
reach any endpoint could be less. Our goal is to describe the distribution of 
the enrollment size of the design for 
planning purposes and to use this distribution to estimate the
success probability.  If all 12 patients are enrolled at once, 
as in the classic 
design, then the sample size is 12. However, in most clinical trials, the 
patients are enrolled sequentially, often with one patient's outcome realized 
before the next one enters the trial. In the present example, observing two 
successful patients allows us reach one endpoint so the sample required 
could be as small as two. So, 12 might not be necessary. Similarly 11 
observed treatment failures also ends the stage. This sampling mechanism, in 
which the experiment ends as soon as any of the endpoints is reached, is 
call {\em curtailed sampling}. Under curtailed sampling the range of the 
sample size is between two and 12.

We assume each of the patient outcomes can be modeled as independent, 
identically distributed Bernoulli($p$) random variables. The realization of 
the trial is a sequence of these random variables that stops when either a 
specified number of success or failures is reached. In the 
previous example suppose two successes were reached after enrolling 10 
patients (one in the third step and one at the $10^{th}$). The sample
path is illustrated 
in Fig.~\ref{fig:kane_viz}. The vertical axis denotes the number of 
successful outcomes. The horizontal axis counts the number of patients that 
have been enrolled. The horizontal and vertical boundaries represent 
endpoints for the trial.

@@

Let $\,b_1, b_2, \ldots \,$ denote a sequence of independent, identically 
distributed, Bernoulli random variables with $\mathbb{E}I_{\{b_i=1\}}=p$ and
$\mathbb{E}I_{\{b_i = 0\}} = 1-p$, for 
probability parameter $0\leq p \leq 1$. In the clinical trial setting 
$\,b_i = 1$ corresponds to a successful outcome.  Let $s$ and $t$ be 
positive integers.  Define the stopped negative binomial (SNB) random 
variable $Y$ as the smallest 
integer value such that $\,\{b_1, \ldots , b_Y\}\,$ contains {\em either} 
$\,s\,$ successes {\em or} $\,t\,$ failures. That is, either 
$\sum_i^Y b_i s$ or else $\sum_i^Y 1-b_i = t$.

The distribution of $\,Y\,$ has support on integer values in the range 
\begin{equation*}               
     \min(s,t) \leq \; Y \;\leq s+t-1  \label{range.y.eq}
\end{equation*}
and it is distributed as
\begin{equation} \label{eqn:pmf}
\mathbb{E}\  I_{\{Y=k\}} = S(k, p, s) I_{\{s \leq k\}} + 
  S(k, 1-p, t) I_{\{ t \leq k \}}
\end{equation}
where $I_{\{f\}}$ is one if $f$ is true and zero otherwise and
\begin{equation} \label{eqn:N}
S(k, p, s) = {k-1 \choose s-1} p^s (1-p)^{k-s} 
\end{equation}
$S$ is the negative binomial probability 
In~\ref{eqn:pmf} we can see that the mass at 
any point on the support of the distribution may come from $S(p)$; $S(1-p)$;
or both of their probabilities.

@@

To show the result in Eqn.~\ref{eqn:pmf} consider the 
process $\mathbf{X} = \left\{X(k) : k = 0,1,... \right\}$ 
with $X(0)=0$ and
\begin{equation*} \label{eqn:proc}
X_{k+1} = X_k + b_{k+1} I_{\{ k-t < X_k < s\}}.
\end{equation*}
The process can be conceptualized as a series of coin 
flips that stops when either $s$ successes or $t$ failures are reached. 
At each step a coin is flipped. If it is heads, the process advances 
one diagonally in the 
positive horizontal and vertical direction in Fig.~\ref{fig:kane_viz}. 
Otherwise, it advances in the positive horizontal direction only. When the 
process stops either $X_k = s$ or $X_k = k-t$.

\begin{prop}
The distribution of the stopping time 
$\argminl_k \left\{ X_k \geq s \cup X_k \leq k-t \right\}$ 
is equal to the SNB and is given in (\ref{eqn:pmf}). 
\end{prop}
\begin{proof}
%The proof will proceed in two parts. First, a combinatorial justification 
%will be given for the probability mass value on each element of the support. 
%Second, it will be shown that the sum of the masses over the support sums to 
%one.

The probability that a given realization of $\mathbf{X}$ reaches $s$ at 
the $k$th outcome is the probability that, at time $k-1$ there are $s-1$ 
successful outcomes and $k-s$ unsuccessful outcomes multiplied by 
the probability of a 
success at time $k$ (\ref{eqn:N}). The probability $X_k = k-t$ 
is the probability that, at outcome $k-1$ there are $k-t$ successful outcomes 
and $t-1$ unsuccessful outcomes multiplied by the probability of an 
unsuccessful outcome at time $k$.  Finally, we need to show that 
the sum
\begin{align} \label{eqn:sum_proof}
R &= \sum_{k=s}^{s+t-1} S(k, p, s) + \sum_{k=t}^{s+t-1} S(k, 1-p, t) \\
  &= \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s (1-p)^{k-s} + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} (1-p)^t
\end{align}
is equal to one.

Substitute $i=k-s$ in the first summation and
$j=k-t$ in the second. 
%\begin{equation*}
%{j+s-1 \choose s-1} = {j+s-1 \choose j}
%\end{equation*}
%Then %(\ref{eqn:sum_proof}) becomes:
%\begin{equation*}
%R = \sum_{i=0}^{t-1} {i+s-1 \choose s-1} p^s (1-p)^i +
%\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
%\end{equation*}
Then $R$ can be written
as the cumulative distribution function of two 
negative binomial distributions
\begin{equation} \label{eqn:transformed_sum}
R = \sum_{i=0}^{t-1} {i+s-1 \choose i} p^s (1-p)^i \; + \;
\sum_{j=0}^{s-1} {j+t-1 \choose j} p^j (1-p)^t.
\end{equation}

Let $\mathcal{I}_p(s, t)$ be the {\em regularized incomplete beta function}, which is 
also one minus the c.d.f. of the negative binomial distribution. This
function satisfies $\mathcal{I}_p(s, t) = 1-\mathcal{I}_{1-p}(t, s)$.  
Then it can be proven
that \ref{eqn:pmf} is a valid probability mass by showing that the
probability mass function sums to one.
\begin{align*}
R = \sum_{i=0}^{t-1} &{i+s-1 \choose i} p^s (1-p)^i +
\sum_{j=0}^{s-1}  {j+t-1 \choose j} p^j  (1-p)^t \\
   &= 1-\mathcal{I}_p(s, t) + 1 - \mathcal{I}_{1-p}(t, s) \\
   &= 1. 
\end{align*}
\end{proof}

\begin{prop} Let $S$ be distributed SNB with parameters $p$, $s$, and $t$.
Then the moment generating function (MGF) of $S$ is
\begin{equation} \label{eqn:mgf}
\mathbb{E}~e^{xS} = \frac{p e^x}{1 - qe^x} \mathcal{I}_{qe^x} (s, t) + 
  \frac{qe^x}{1-pe^x} \mathcal{I}_{pe^x}(s, t)
\end{equation}
for $q = 1-p$ when $x \leq -\log(q)$ and $x \leq -\log(p)$.
\end{prop}
\begin{proof}
By definition, the MGF of the SNB is:
\begin{equation*}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1} {k-1 \choose s-1} p^s q^{k-s} e^{kx} 
  + \sum_{k=t}^{s+t-1} {k-1 \choose k-t} p^{k-t} q^t e^{kx}
\end{equation*}
and this can be rewritten as:
\begin{equation} \label{eqn:first_sum}
\mathbb{E}~e^{xS} = \sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} 
  + \sum_{k=t}^{s+t-1}{k-1 \choose k-t} (qe^x)^t (pe^x)^{k-t}.
\end{equation}
Taking the first summation in Equation \ref{eqn:first_sum}:
\begin{align*}
\sum_{k=s}^{s+t-1}{k-1 \choose s-1} (pe)^{sx} (qe^x)^{k-s} &= 
  \left(\frac{pe^x}{1 - qe^x}\right)^s \ \ \sum_{k=s}^{s+t-1} {k-1 \choose s-1} 
    (qe^x)^{k-s} (1-qe^x)^s) \\
  &= \left(\frac{pe^x}{1 - qe^x}\right)^s \mathcal{I}_{qe^x}(s, t).
\end{align*}
Since the incomplete beta function has support on zero to one 
$qe^x \leq 1$. This implies that $x \leq -\log(q)$.

@@

Let us assume that $p$ has a Beta distribution, with constant prior 
parameters $\alpha$ and $\beta$. Then the closed form 
posterior distribution of the SNB is
\begin{align}
f \left(k | \mathbf{b}, \alpha, \beta \right) &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
  I_{\{s \leq k \leq s+k-1\}} + \nonumber \\
  & {k-1 \choose k-t} 
  \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
  I_{\{t \leq k \leq s+k-1\}}
\end{align}
where $B$ is the Beta function and $\mathbf{b}$ is the realization 
of Bernoulli data.

@@

\begin{prop}
The posterior PMF of the Stopped Negative Binomial distribution with a Beta($\alpha$, $\beta$) prior is:
\begin{align} \label{eqn:posterior}
f(k | s, t, \alpha, \beta) &= 
  {k-1 \choose s-1} \frac{B\left(\alpha+s, k-s+\beta \right)}{B(\alpha, \beta)} 
    \ I_{\{s \leq k \leq s+k-1\}} \ + \nonumber \\
  & {k-1 \choose k-t} 
    \frac{B\left(\alpha + k - t, t+\beta\right)}{B(\alpha, \beta)} 
    \ I_{\{t \leq k \leq s+k-1\}}
\end{align}
where $B$ is the Beta function.
\end{prop}
\begin{proof}
For notational simplicity, assume that $s,t \leq k \leq s+t-1$. When this is not the case appropriate terms should be removed as dictated by the indicator functions.
\begin{align*}
f(k | s, t, \alpha, \beta) = \frac{1}{B(\alpha, \beta)} & \int_0^1 {k-1 \choose s-1} p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} + \\
 & {k-1 \choose k-t} p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp \\
= \frac{1}{B(\alpha, \beta)}  {k-1 \choose s-1} & \int_0^1  p^{\alpha +s -1} \left(1-p\right)^{k-s+\beta-1} dp + \\
 & \frac{1}{B(\alpha, \beta)} {k-1 \choose k-t} \int_0^1  p^{k-t+\alpha-1}\left(1-p\right)^{t+b-1} dp
\end{align*}
The result follows by applying the definition of the Beta function to the integral terms.
\end{proof}
